\MakeUppercase{Notes on Layered $H$-Partitions of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-05-01 warmup.tex
Fuck yeah!

\documentclass{patmorin}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{pat}
\usepackage{paralist}
%\usepackage{hyperref}

@@

\DeclareMathOperator{\dist}{dist}

@@

>>>>

\documentclass{patmorin}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{pat}
\usepackage{paralist}

\usepackage{tipa}
\usepackage{upgreek}
\usepackage{rotating}
\newcommand{\V}{\rotatebox[origin=c]{180}{$V$}}
\newcommand{\Y}{\rotatebox[origin=c]{180}{$Y$}}
%\usepackage{hyperref}

@@

\DeclareMathOperator{\dist}{dist}
\DeclareMathOperator{\depth}{depth}

@@

\section{$k$-Planar}

  We make use of the fact that, if $G$ is a $k$-plane graph and $G^+$ is the plane graph obtained by adding a dummy vertex at each crossing in $G$, then $\dist_{G^+}(x,y)\le k\dist_G(x,y)$ for any $x,y\in V(G)$.

\begin{lem}\lemlabel{k-induction} The setup:
  \begin{compactenum}
    \item Let $G_0$ be a $k$-plane graph.
    \item Let $G=(V(G_0), E(G))$ be some subgraph of $G_0$ (so $E(G)\subseteq E(G_0)$).
    \item Let $G^+$ be the plane graph obtained by adding a vertex at each crossing point in $G_0$.
    \item Let $T$ be a spanning tree of $G^+$ rooted at some vertex $r$.
    \item For each integer $i\ge 0$, let $V^+_i=\{v\in V(G^+):\dist_{G^+}(r,v)=i\}$, and let $V_i=V(G)\cap V^+_i$.
    \item Let $C$ be a cycle in $G^+$ with $r$ in the exterior of $C$ and such that
    \begin{compactenum} 
      \item No edge of $C$ is crossed by any edge of $G$; and
      \item $V(C)$ can be partitioned into $P_1,\ldots,P_k$, $k\le 3$ such that for each $i\in\{1,\ldots,k\}$,
      \begin{compactenum}
        \item $C[P_i]$ is a path; and
        \item $|P_i\cap V_j| \le 24k+4$ for all integers $j\ge 0$.
      \end{compactenum}
    \end{compactenum}
    \item Let $N$ and $N^+$ be the subgraphs of $G$ and $G^+$ consisting only of those edges and vertices contained in $C$ or the interior of $C$.
  \end{compactenum}
  Then $N$ has an $H$-partition $(B_x : x\in V(H))$ such that
  \begin{compactenum}
    \item for every integer $j\ge 0$, and all $x\in V(H)$, $|B_x\cap V_j|\le 24k+4$; 
    \item for each $i\in\{1,\ldots,k\}$, there exists some $x_i\in V(H)$ such that $P_i=B_{x_i}$; and
    \item $H$ has a tree decomposition whose largest bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
  \end{compactenum}
\end{lem}

\begin{proof}
  The proof is by induction on the number of vertices of $N^+$.

  If $k=3$, set $R_i=P_i$ for each $i\in\{1,2,3\}$.  Otherwise, as before, split $P_1,\ldots,P_k$ to partition $V(C)$ into three sets $R_1$, $R_2$, and $R_3$ such that each $C[R_i]$ is a path and each $R_i$ contains vertices from at most one of $P_1,\ldots,P_k$.

  Next, we 3-colour $V(N^+)$ so that we can apply Sperner's Lemma.
  For each $i\in\{1,2,3\}$ and $v\in R_i$, assign the colour $i$ to $v$.
  For each $v\in V(N^+)$, consider the path $P_v$ that consists of the subpath of the $v$ to $r$ path in $T$ that stops at the first vertex $v'\in V(C)$. Assign $v$ to have the same colour as $v'$.

  Add edges to $N^+$ to obtain a near-triangulation $G^\Delta$.  
  Sperner's Lemma ensures that $G^\Delta$ contains a triangular face $\tau=v_1v_2v_3$ whose three vertices are assigned different colours. For each $i\in\{1,2,3\}$, let $Q_i^+=P_{v_i}^+$ be the path in $T$ from $v_i$ to $v_i'$.  Let $Y^+$ denote the subgraph of $N^+$ consisting of vertices and edges of $Q_1^+$, $Q_2^+$, $Q_3^+$, and $\tau$.

  Define a vertex set $X\subset V(G)$ as follows:  The set $X$ contains $V(Y^+)\cap V(G)\setminus V(C)$.  Next, for each vertex $v^+\in V(Y^+)\setminus V(G)$, we consider the two edges $vw,xy\in E(G_0)$ whose intersection is $v^+$.  Next we set
  \[   X_{vw} = \begin{cases} \{v,w\} & \text{if $vw\in E(G)$} \\
           \emptyset & \text{if $vw\not\in E(G)$,}
         \end{cases}
  \]
  \[   X_{xy} = \begin{cases} \{x,y\} & \text{if $xy\in E(G)$} \\
           \emptyset & \text{if $xy\not\in E(G)$} 
         \end{cases}
  \]
  and add $(X_{vw}\cup X_{xy})\setminus V(C)$ to $X$.

  We claim that, for each integer $i\ge 0$, $|V_i\cap X|\le 24k+4$.  To see why this is so, first note that $|V_i^+\cap V(Y^+)|\le 3$ since $V_i^+$ contains at most one vertex from each of $Q_1^+$, $Q_2^+$, and $Q_3^+$. For any vertex $v\in X$, there exists a vertex $v^+\in V(Y^+)$ such that $\dist_{G'}(v,v^+)\le k$.  Turning this around, we can assign each vertex $v\in X$ to a vertex $v^+\in V(Y^+)$ in such a way that each vertex of $Y^+$ is assigned at most 4 vertices of $X$ (each dummy vertex $v^+$ is assigned the four endpoints of the crossing pair of edges that generated $v^+$).  Since any vertex $v\in X\cap V_i$ is assigned to one of the at most $6(k+1)$ vertices in $X\cap (V(Y^+)\cap( V_{i-k}^+\cup\cdots\cup V_{i+k}^+))$, we have $|X\cap V_i|\le 24k+4$.

  Let $S^+$ denote the subgraph of $N^+$ consisting of the vertices and edges of $Y^+$ and the vertices and edges of $C$.  The graph $S^+$ is planar. It contains one triangular face $\tau$ and up to three additional bounded faces $F_1$, $F_2$, and $F_3$, where each $F_i$ does not have any vertices in common with $P_i$.

  For each $i\in\{1,2,3\}$, let $Z_i$ be the subset of $V(G)$ that is contained in the interior of $F_i$.  We claim that $N-X-V(C)$ does not contain any edge $vw$ with $v\in Z_i$ and $w\in Z_j$ for $i\neq j$.  Indeed, any such edge would contain a vertex of $V(Y^+)\setminus V(G)$ in its interior and therefore each of $v,w\in X\cup V(C)$.

  The set $X$ will become an element $B_x$ in our $H$-partition of $G$.  All that remains is to apply induction on three subgraphs.  For each $i\in\{1,\ldots,3\}$, let $N_i^+$ denote the subgraph of $G^+$ containing the vertices and edges of $G^+$ that are contained in the closure of $F_i$.  
  Let $G_i^-$ be the subgraph of $G$ obtained by removing the edges of $G$ that cross the boundary of $F_i$.  
  Let $N_i$ denote the subgraph of $G_i^-$ containing the vertices and edges of $G_i^-$ that are contained in the closure of $F_i$.  Note that $G_i^-$ is a subgraph of $G_0$ and that the number of vertices of $N_i^+$ is less than the number of vertices of $N^+$ since $N_i^+$ does not contain one of the vertices of $\tau$.  Therefore, we can apply induction to obtain the desired $H$-partition of $N_i$ using the original graphs $G_0$ and $G^+$, the subgraph $G_i^-$ in place of $G$, the cycle $F_i$ in place of $C$, and partition of $V(F_i)$ into $V(F_i)\cap X$, $V(F_i)\cap P_{i-1}$, and $V(F_i)\cap P_{i+1}$.

  The remainder of the proof is as before. We use $X$ as an element of our $H$-partition. The root bag in the tree decomposition contains the vertices obtained by contracting each of $P_1,\ldots,P_k$ and $X$.
\end{proof}

\begin{thm}\thmlabel{1-plane}
  Every $k$-plane graph $G$ has a $(3, 24k^2+4k)$-partition.
\end{thm}

\begin{proof}
  The same as before except that the layered width $24k+4$ from \lemref{k-induction} becomes $24k^2+4k$ because \lemref{k-induction} uses a layering of $G^+$ and distances in $G^+$ can be a factor of k larger than in $G$.
\end{proof}

% Let $G$ be a $k$-planar graph with a distinguished vertex $r$. For each integer $i\ge 0$, let $V_i=\{v\in V(G):\dist_G(r,v)=i\}$ and let $V_{\le i}=\bigcup_{j=0}^i V_j$.  We will use distances to $r$ frequently, so for each $v\in V(G)$, we let $d_v=\dist_G(r,v)$ (so that $v\in V_{d_v}$). We say that a subgraph $P$ of $G$ has layered width at most $\ell$ if $|V_i\cap V(P)|\le \ell$ for each integer $i\ge 0$.
% 
% Let $P_1,\ldots,P_3$ be three vertex-disjoint subgraphs of $G$, each of which has layered width at most $\ell$ and none of which contains $r$. Let $B$ be the union of $P_1,\ldots,P_3$.  Let $N$ be a component of $G-V(B)$ that does not contain $r$.  Let $G'$ be the subgraph of $G$ consisting of edges and vertices of $N$, edges and vertices of $B$, and all edges $vw\in E(G)$ with $v\in V(N)$ and $w\in V(B)$.
% 
% \begin{lem}
%   There exists a vertex $v\in V(G')$ and three vertex-disjoint paths $Q_1$, $Q_2$, and $Q_3$ where each $Q_i$ begins at $v$, has all of its internal vertices in $V(N)$, ends at a vertex in $V(P_i)$ and has layered width at most 10. 
% \end{lem}
% 
% \begin{proof}
%   For simplicity, assume that every vertex of $P_i$ is adjacent to at least one vertex of $V(N)$ since we can ignore vertices of $P_i$ for which this is not the case.  
% 
%   For each integer $r\ge 0$, consider the induced subgraph $G_r'=G'[V(B)\cup V_{\le r}]$.  Let $r$ be the smallest integer such that $G_{r}'$ has at least one component $A$ that contains vertices from each of $P_1$, $P_2$, and $P_3$.  Select any vertex $w\in V(A)\setminus V_{<r}$. Suppose, for now that $w\not\in V(B)$.  Then, since no component of $V_{r-1}$ contains vertice from each of $P_1$, $P_2$, and $P_3$, it must be the case that $G_{r-1}'$ has components $A_1$, $A_2$, and $A_3$ where $w$ is adjacent to some vertex in each $A_i$, each $A_i$ contains a vertex of $P_i$, $A_1\not\in\{A_2,A_3\}$ and $A_1$ does not contain a vertex of $P_3$.  (Note that we may have $A_2=A_3$ and we may have $A_1$ containig vertices of $P_2$.) 
% 
%   If $A_2\neq A_3$, then we are done.  We set $v=w$ and, for each $i\in \{1,\ldots,3\}$ we take a path $Q_i$ from $v$ to $V(P_i)$ whose $j$th vertex is in $V_{r-j+1}\cap V(A_i)$.  These three paths are disjoint since $V(Q_i)\subseteq V(A_i)$ for each $i\in \{1,\ldots,3\}$.
% 
%   If $A_2=A_3$, there is still a little more work to do.  In this case we will still use the path $Q_1$ from $w$ to $V(P_1)$.  Now, $w$ is adjacent to a vertex $w_1$ in $A_2=A_3$ so we extend
% 
% 
% 
% 
% 
%    where (without loss of generality) $V(A_1)\cap V(P_1)\neq \emptyset$, $V(A_1)\cap (V(P_2)\cup V(P_3))=\emptyset$, $V(A_2)\cap V(P_2)\neq\emptyset$, $V(A_2)\cap V(P_1)=\emptyset$ and $w$ is adjacent to a vertex $w_i\in A_i$ for each $i\in\{1,2\}$.
% 
%   If 
% 
% 
% 
% 
% 
%    contains vertices of $P_1$, $A_2$ contains vertices of $P_2$ and (possibly )
% 
% 
% 
% 
%   We claim that $v$ has the desired property.  Indeed
% 
% \end{proof}
% 
% 
% 
% 
% 
% \section{A Metric Alternative To Sperner's Lemma}
% 
% I'm interested in having something more robust and less magical than Sperner's Lemma. Here's a first attempt.
% 
% Let $G$ be a planar triangulation with some distinguished vertex $r$. 
% 
% 
% 
% 
% A \emph{$t$-pod} (with respect to $r$) consists of vertex-disjoint paths $Q_1,\ldots,Q_t$, where each $Q_i$ a prefix of some $v_i\leadsto r$ path in $G$ and $v_1,\ldots,v_t$ form a clique in $G$.  We call $v_1,\ldots,v_t$ the \emph{central vertices} of the tripod and we call $Q_1,\ldots,Q_t$ the \emph{legs} of the tripod.  The final vertex in each path $Q_i$ is called \emph{foot} if the tripod.  
% 
% A $t$-pod $\Y$ with $t\le 3$, is called a \emph{tripod}.  A $t$-pod $\V$ with $t\le 2$ is called a \emph{bipod}.  A $t$-pod $I$ with $t\le 1$ is called a \emph{monopod}.
% 
% For a $t$-pod $Y$ with central vertices $v_1,\ldots,v_t$, we define the \emph{depth} of $Y$ as 
% \[
%     \depth(Y)=\max\{d_G(r,v):v\in V(T)\} 
%         = \max\{\dist_G(r, v_i): i\in\{1,\ldots,t\}\} \enspace .
% \]
% Observe that, since $v_1,\ldots,v_t$ form a clique in $G$, $\depth(Y)-1\le \dist_G(r, v_i) \le \depth(Y)$.
% 
% Let $C$ be a cycle in $G$ and $P_1,\ldots,P_a$, $a\le 3$ be a partition of $V(C)$ such that $C[P_i]$ is a path and $C[P_i]$ is also a bipod $\V_i$ in $G$. Let $Q_{i,1}$ and (possibly) $Q_{i,2}$ be the legs of $\V_i$ where we use the convention that, for each $i\in\{1,\ldots,a\}$, the last vertex of $Q_{i,2}$ and the first vertex of $Q_{i+1,1}$ are adjacent in $C$.  Let $v_{i,j}$ be first vertex of $Q_{i,j}$ and let $v_{i,j}'$ be the last vertex of $Q_{i,j}$.  With these definitions, $v_{i,j}$ is a central vertex of $\V_i$ and $v_{i,j}'$ is a foot of $\V_{i}$.
% 
% Note that, for each $i\in\{1,\ldots,a\}$, let $d_i = dist_{G}(r,\{v_{i,2},v_{i+1,1}'\})$.  
% 
% 
% 
% Let $N$ be the near-triangulation consisting of all edges and vertices of $G$ that are on or in the ineterior of $C$.  
%
